---
title: "Temperature 12k Composites — Reconstruction Validation Diagnostics"
subtitle: "Per-method + consensus GMST vs published Kaufman et al. (2020)"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 8
    fig_height: 4.2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      dev = "png", dpi = 120)
# Run relative to the repo root (this file lives in reproduction/).
knitr::opts_knit$set(root.dir = normalizePath(".."))
library(ggplot2)
```

```{r helpers}
# ---------------------------------------------------------------------------
# Inputs (repo-relative):
#   results/methods/<m>_global.csv   binAges + ensemble-member columns  (mine)
#   results/reconstruction.csv       consensus mean / lo_95 / hi_95
#   reference_data/published/<m>_published.csv   age_bp, median, lo, hi
# Both curves are anchored to 0 at 100 BP, matching reproduction/ci/cmp.py.
# ---------------------------------------------------------------------------
METHODS <- c("scc", "dcc", "gam", "cps", "paico")
LABELS  <- c(scc = "SCC", dcc = "DCC", gam = "GAM", cps = "CPS",
             paico = "PaiCo", consensus = "Consensus")
COLORS  <- c(scc = "#4E79A7", dcc = "#59A14F", gam = "#B07AA1",
             cps = "#1170AA", paico = "#E15759", consensus = "#111111")

anchor <- function(age, y, at = 100) {
  o <- order(age); y - approx(age[o], y[o], xout = at, rule = 2)$y
}

# My reconstruction: median + 5-95% band for a method (or the consensus).
mine_curve <- function(method) {
  if (method == "consensus") {
    d <- read.csv("results/reconstruction.csv", check.names = FALSE)
    age <- d$age_bp
    return(list(age = age, med = d$mean, lo = d$lo_95, hi = d$hi_95))
  }
  d <- read.csv(sprintf("results/methods/%s_global.csv", method), check.names = FALSE)
  age <- d$binAges
  ens <- as.matrix(d[, setdiff(names(d), "binAges"), drop = FALSE])
  list(age = age,
       med = apply(ens, 1, median, na.rm = TRUE),
       lo  = apply(ens, 1, quantile, 0.05, na.rm = TRUE),
       hi  = apply(ens, 1, quantile, 0.95, na.rm = TRUE))
}

# Published (NOAA-archived Kaufman 2020) median + lo/hi for a method.
pub_curve <- function(method) {
  f <- sprintf("reference_data/published/%s_published.csv", method)
  if (!file.exists(f)) return(NULL)
  d <- read.csv(f, check.names = FALSE)
  list(age = d$age_bp, med = d$median, lo = d$lo, hi = d$hi)
}

wmean <- function(age, y, lo, hi) mean(y[age >= lo & age <= hi], na.rm = TRUE)

# Metric set matching cmp.py (anchored at 100 BP).
metrics <- function(method) {
  m <- mine_curve(method); p <- pub_curve(method)
  if (is.null(p)) return(NULL)
  age <- m$age; o <- order(p$age)
  pm  <- approx(p$age[o], p$med[o], xout = age, rule = 2)$y
  plo <- approx(p$age[o], p$lo[o],  xout = age, rule = 2)$y
  phi <- approx(p$age[o], p$hi[o],  xout = age, rule = 2)$y
  me <- anchor(age, m$med); pu <- anchor(age, pm)
  ok <- is.finite(me) & is.finite(pu)
  data.frame(
    method = LABELS[[method]],
    r      = cor(me[ok], pu[ok]),
    bias   = mean(me - pu, na.rm = TRUE),
    RMSE   = sqrt(mean((me - pu)^2, na.rm = TRUE)),
    maxD   = max(abs(me - pu), na.rm = TRUE),
    amp    = sd(me, na.rm = TRUE) / sd(pu, na.rm = TRUE),
    midHol = wmean(age, me, 5500, 6500),
    `12ka` = wmean(age, me, 11500, 12000),
    spread = mean(m$hi - m$lo, na.rm = TRUE) / mean(phi - plo, na.rm = TRUE),
    check.names = FALSE)
}

# One overlay panel: my median + 5-95% ribbon vs published median + lo/hi ribbon.
plot_method <- function(method) {
  m <- mine_curve(method); p <- pub_curve(method); col <- COLORS[[method]]
  age <- m$age; o <- order(p$age)
  pm  <- approx(p$age[o], p$med[o], xout = age, rule = 2)$y
  plo <- approx(p$age[o], p$lo[o],  xout = age, rule = 2)$y
  phi <- approx(p$age[o], p$hi[o],  xout = age, rule = 2)$y
  a0 <- function(y) anchor(age, y)
  df <- data.frame(kyr = age / 1000,
                   me = a0(m$med), mlo = a0(m$lo), mhi = a0(m$hi),
                   pu = a0(pm), plo = a0(plo), phi = a0(phi))
  ggplot(df, aes(kyr)) +
    geom_ribbon(aes(ymin = plo, ymax = phi), fill = "grey55", alpha = 0.25) +
    geom_ribbon(aes(ymin = mlo, ymax = mhi), fill = col, alpha = 0.22) +
    geom_line(aes(y = pu), colour = "grey25", linewidth = 0.7, linetype = "22") +
    geom_line(aes(y = me), colour = col, linewidth = 1.0) +
    scale_x_reverse(breaks = seq(0, 12, 2)) +
    labs(x = "Age (ka BP)", y = expression("Temperature anomaly ("*degree*"C)"),
         title = sprintf("%s — mine vs published (anchored 100 BP)", LABELS[[method]]),
         caption = "solid = mine (median, 5–95% band); dashed grey = published (median, band)") +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
}
```

## Summary

All curves are anchored to 0 at 100 BP and scored exactly as `reproduction/ci/cmp.py`
does. Data are the real-ensemble (v1.0.0) per-method global ensembles in
`results/methods/` and the NOAA-archived published curves in
`reference_data/published/`.

```{r table}
tab <- do.call(rbind, lapply(c(METHODS, "consensus"), metrics))
knitr::kable(tab, digits = 3, row.names = FALSE,
             caption = "Validation metrics vs published (target: maxD 0, bias 0, amp 1, spread 1).")
```

## Per-method overlays {.tabset}

```{r per-method, results="asis", fig.height=4.2}
for (mth in METHODS) {
  cat(sprintf("\n\n### %s\n\n", LABELS[[mth]]))
  print(plot_method(mth))
  mt <- metrics(mth)
  cat(sprintf("\n\n*maxD %.3f · bias %+.3f · amp %.2f · spread %.2f · midHol %.2f (pub via published)*\n",
              mt$maxD, mt$bias, mt$amp, mt$spread, mt$midHol))
}
```

## Consensus

The multi-method consensus (all five methods' members pooled) vs the published
consensus.

```{r consensus, fig.height=4.2}
plot_method("consensus")
```

## Combined overlay

All five method medians + the consensus, each anchored at 100 BP, over the
published consensus band.

```{r combined, fig.height=5}
p <- pub_curve("consensus")
age <- mine_curve("consensus")$age; o <- order(p$age)
band <- data.frame(kyr = age/1000,
                   lo = anchor(age, approx(p$age[o], p$lo[o],  xout = age, rule = 2)$y),
                   hi = anchor(age, approx(p$age[o], p$hi[o],  xout = age, rule = 2)$y))
lines <- do.call(rbind, lapply(c(METHODS, "consensus"), function(mth) {
  m <- mine_curve(mth)
  data.frame(kyr = m$age/1000, y = anchor(m$age, m$med), method = LABELS[[mth]])
}))
lines$method <- factor(lines$method, levels = unname(LABELS[c(METHODS, "consensus")]))
ggplot() +
  geom_ribbon(data = band, aes(kyr, ymin = lo, ymax = hi), fill = "grey60", alpha = 0.3) +
  geom_line(data = lines, aes(kyr, y, colour = method,
            linewidth = method == "Consensus")) +
  scale_colour_manual(values = setNames(unname(COLORS[c(METHODS,"consensus")]),
                                         unname(LABELS[c(METHODS,"consensus")]))) +
  scale_linewidth_manual(values = c(`FALSE` = 0.7, `TRUE` = 1.6), guide = "none") +
  scale_x_reverse(breaks = seq(0, 12, 2)) +
  labs(x = "Age (ka BP)", y = expression("Temperature anomaly ("*degree*"C)"),
       colour = NULL, title = "All methods + consensus vs published band",
       caption = "grey band = published consensus 5–95%") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom",
        panel.grid.minor = element_blank())
```

---
*Generated from `reproduction/diagnostics.Rmd`. Regenerate after a new
reconstruction with `rmarkdown::render("reproduction/diagnostics.Rmd")`.*
